The Current State of R Tools for Nonlinear Least Squares Modeling

### 2023-07-15 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
notes from AB

---
title: "The Current State of R Tools for Nonlinear Least Squares Modeling"
author:
- John C. Nash \thanks{ retired professor, Telfer School of Management, University
  of Ottawa}
- Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian
  Institute of Technology, Kanpur}
date: "2023-3-2"
output:
  pdf_document:
    keep_tex: yes
    toc: no
  word_document:
    toc: no
bibliography: ./StateOfRToolsForNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

@@

Let us consider there are $n$ parameters and $m$ residuals. 
The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ g = 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  H_{i,j}/2 = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta  = -g $$
Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ H \delta / 2 = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ H / 2 \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by not forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods. On the other hand, when $m >> n$,
the matrix $J' J$ uses much less storage than $J$.

@@

There are problems where use of R will be slower. For example, a
"large" case of the Penalty 1 test problem of @More1981TUO will take
noticeable time to solve. Problems in large numbers of parameters
create large computational matrices for the Jacobian or its inner product,
which Gauss-Newton methods create and process. 
This is confirmed by comparing `nlsr::nlfb()` and `minpack.lm::nls.lm()` on
the Penalty 1 problem of various sizes, where the latter, which is 
compiled Fortran internally, was moderately faster. However, the method
"Rcgmin" from package `optimx` (@p-optimx) proved to be much 
faster than either, despite being entirely written in R, and is a much more
reliable conjugate gradient minimizer than method "CG" of `optim()`. One
of us created both, but now deprecates the latter. 
Timings of CG methods are more sensitive to small differences in
setup and arithmetic than Gauss-Newton methods, and the pattern of
such timings varies much less regularly with problem size for CG methods.

>>>>

---
title: "The Current State of R Tools for Nonlinear Least Squares Modeling"
author:
- John C. Nash \thanks{ retired professor, Telfer School of Management, University
  of Ottawa}
- Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian
  Institute of Technology, Kanpur}
date: "2023-3-2"
output: 
  pdf_document:
    pandoc_args: --listings
    includes:
      in_header: preamble.tex
  word_document:
    toc: no
bibliography: ./StateOfRToolsForNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

@@

Let us consider there are $n$ parameters and $m$ residuals. 
The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ g = 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  H_{i,j}/2 = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta  = -g $$
Given a set of parameters $p$, which provide $H$ and $g$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ H \delta / 2 = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ H / 2 \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by not forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods. On the other hand, when $m >> n$,
the matrix $J' J$ uses much less storage than $J$.

@@

There are problems where use of R will be slower. For example, a
"large" case of the Penalty 1 test problem of @More1981TUO will take
noticeable time to solve. Problems in large numbers of parameters
create large computational matrices for the Jacobian or its inner product,
which Gauss-Newton methods create and process. 
This is confirmed by comparing `nlsr::nlfb()` and `minpack.lm::nls.lm()` on
the Penalty 1 problem of various sizes, where the latter, which uses 
compiled Fortran internally, was moderately faster. However, the method
"Rcgmin" from package `optimx` (@p-optimx) proved to be much 
faster than either, despite being entirely written in R, and is a much more
reliable conjugate gradient minimizer than method "CG" of `optim()`. One
of us created both, but now deprecates the latter. 
Timings of CG methods are more sensitive to small differences in
setup and arithmetic than Gauss-Newton methods, and the pattern of
such timings varies much less regularly with problem size for CG methods.
